---
title: "A07- ggplot"
author: "Vydeepthi Dhulipalla"
date: "June 8,2023"
output: html_notebook
---

## Bicycle Sharing

Capital BikeShare is a bicycle-sharing system in Washington, D.C. At any of about 400 stations, a registered user can unlock and check out a bicycle. After use, the bike can be returned to the same station or any of the other stations.

Such sharing systems require careful management. There need to be enough bikes at each station to satisfy the demand, and enough empty docks at the destination station so that the bikes can be returned. At each station, bikes are checked out and are returned over the course of the day. An imbalance between bikes checked out and bikes returned calls for the system administration to truck bikes from one station to another. This is expensive.

In order to manage the system, and to support a smart-phone app so that users can find out when bikes are available, Capital BikeShare collects real-time data on when and where each bike is checked out or returned, how many bikes and empty docks there are at each station. Capital BikeShare publishes the station-level information in real time. The organization also publishes, at the end of each quarter of the year, the historical record of each bike rental in that time.

You can access the data from the Capital Bikeshare web site. Doing this requires some translation and cleaning, skills that are introduced in Chapter 16. For this project, however, already translated and cleaned data are provided in the form of two data tables:

- `Stations` giving information about location of each of the stations in the system.
- `Trips` is a random subset of 10,000 trips from the full quarterly data, giving the rental history over the last quarter of 2014 (Q4).


```{r}
library(tidyverse)

station_url <- "https://mdbeckman.github.io/dcSupplement/data/DC-Stations.csv"
Stations <- readr::read_csv(station_url)
    
trip_url <- "https://mdbeckman.github.io/dcSupplement/data/Trips-History-Data-2014-Q4-Small.rds"
Trips <- readRDS(gzcon(url(trip_url)))
```


In this activity, you’ll work with just a few variables:

From `Stations`:

- the latitude (`lat`) and longitude (`long`) of the bicycle rental station
- `name`: the station’s name


From Trips: 

- `sstation`: the name of the station where the bicycle was checked out.
- `estation`: the name of the station to which the bicycle was returned.
- `client`: indicates whether the customer is a "regular" user who has paid a yearly membership fee, or a "casual" user who has paid a fee for five-day membership.
- `sdate`: the time and date of check-out
- `edate`: the time and date of return

(Notice that the location and time variables start with an “s” or an “e” to indicate whether the variable is about the **s**tart of a trip or the **e**nd of a trip.)


Time/dates are typically stored as a special kind of number: a **POSIX** date. 

**POSIX** date: A representation of date and time of day that facilitates using dates in the same way as numbers, e.g. to find the time elapsed between two dates. You can use `sdate` and `edate` in the same way that you would use a number. For instance, the figure below shows the distribution of times that bikes were checked out.

```{r}
Trips %>%
  ggplot(aes(x = sdate)) +
  geom_density(fill = "gray", color = NA)
```


## Question 1: How Long?  

Make a box-and-whisker plot, like Figure [18.2 of the Project: Bicycle Sharing](https://dtkaplan.github.io/DataComputingEbook/project-bicycle-sharing.html#fig:box-whisker-bike) showing the distribution of the duration of rental events, broken down by the client type. The duration of the rental can be calculated as `as.numeric(edate - sdate)`. The units will be in either hours, minutes, or seconds. It should not be much trouble for you to figure out which one.

When you make your plot, you will likely find that the axis range is being set by a few outliers. These may be bikes that were forgotten. Arrange your scale to ignore these outliers, or filter them out.


```{r}
### Your code here
# Calculate the duration of rental events
Trips <- Trips %>%   # Update the Trips dataset
  mutate(duration = as.numeric(edate - sdate))

# Filter out outliers
quantile_cutoff <- quantile(Trips$duration, 0.99)  # Calculate the 99th percentile
Trips_filtered <- Trips %>%   # Create a new dataset with filtered data
  filter(duration < quantile_cutoff)

# Create the box-and-whisker plot
ggplot(data = Trips_filtered, aes(x = client, y = duration/60)) +   # Specify the dataset and mapping variables
  geom_boxplot() +   # Add the box-and-whisker plot layer
  ylab("Duration (minutes)") +   # Set the y-axis label
  xlab("Client Type")   # Set the x-axis label


```


## Question 2: When are bikes used?

The `sdate` variable in Trips indicates the date and time of day that the bicycle was checked out of the parking station. `sdate` is stored as a special variable type that includes detailed information about the date and time. 

Often, you will want discrete components of a date, for instance:



|Date Component                         |	Function (`lubridate` package)|
|---------------------------------------|-----------------------------|
| Day of the year (1-365)              	| `lubridate::yday(sdate)`	  | 
| Day of the week (Sunday to Saturday)	| `lubridate::wday(sdate)`	  | 
| Hour of the day		                    | `lubridate::hour(sdate)`	  | 
| Minute in the hour	                  | `lubridate::minute(sdate)`	| 

Create **histograms or density plots** for bike use over the following discrete time periods. For each plot, provide 2-3 sentences describing your observations. 

1. day of the year (1 to 365)
2.  day of the week (Sunday to Saturday)
3.  hour of the day (0 to 24)
4.  minute in the hour (0 to 60)

You can use [Figure 18.3](https://dtkaplan.github.io/DataComputingEbook/project-bicycle-sharing.html#fig:bikes-over-hours) or [Figure 18.4](https://dtkaplan.github.io/DataComputingEbook/project-bicycle-sharing.html#fig:bike-hour-count) as examples.


### Day of Year

```{r}
Trips %>%
  ggplot(aes(x = lubridate::yday(sdate))) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  xlab("Day of the Year") +
  ylab("Count") +
  ggtitle("Bike Use by Day of the Year")


```



```{r}
observation1 <- "The graph shows how often bikes are used at different times of the year. It looks like there aren't as many bike rentals in the winter (between day 300 and day 360) as there are the rest of the year. Around day 200, there seems to be a lot of bike action, which is when the usage seems to peak."
cat(observation1)
```

### Day of Week

```{r}
Trips %>%
  ggplot(aes(x = lubridate::wday(sdate))) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  scale_x_continuous(breaks = 1:7, labels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")) +
  xlab("Day of the Week") +
  ylab("Count") +
  ggtitle("Bike Use by Day of the Week")


```


```{r}
observation2 <- "The graph shows how often people ride bikes on different days of the week. It looks like more people rent bikes Monday through Friday than on the weekends (Saturday and Sunday). This means that most people use the bike-sharing system to get to and from work or school."

cat(observation2)
```



### Hour of Day

```{r}
Trips %>%
  ggplot(aes(x = lubridate::hour(sdate))) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  xlab("Hour of the Day") +
  ylab("Count") +
  ggtitle("Bike Use by Hour of the Day")

```


My observations here..... 

```{r}
observation3 <- "The histogram shows how many bike hires are made at different times of the day. It has a two-peak schedule, with the most people riding bikes during rush hour in the morning (around 8–9 am) and in the evening (around 5–6 pm). This shows that people who commute to and from work use the bike-sharing system a lot."
cat(observation3)
```


### Minute of Hour

```{r}
Trips %>%
  ggplot(aes(x = lubridate::minute(sdate))) +
  geom_density(fill = "lightblue", color = "black") +
  xlab("Minute in the Hour") +
  ylab("Density") +
  ggtitle("Bike Use by Minute in the Hour")

```


```{r}
observation4 <- "The density plot shows how many bike rentals there are during each hour. It shows that bike use is pretty even across the minutes in an hour, which means that no one minute is much more popular than the others. This means that there are an equal number of bike hires every hour."
cat(observation4)
```

## Question 3: Grouping

Group the bike rentals by three variables: hour of the day, day of the week, and client type. Find the total number of events in each grouping and plot this count versus hour. Use the group aesthetic to represent one of the other variables and faceting to represent the other. Comment on any general trends you see.

```{r}
# Load required packages
library(dplyr)
library(ggplot2)

# Extract hour of the day from sdate
Trips$hour <- lubridate::hour(Trips$sdate)

# Extract day of the week from sdate
Trips$day_of_week <- lubridate::wday(Trips$sdate, label = TRUE)

# Group the data by hour, day of the week, and client type and calculate the count
grouped_data <- Trips %>%
  group_by(hour, day_of_week, client) %>%
  summarise(count = n())

# Plot the count versus hour, faceted by day of the week and colored by client type
ggplot(grouped_data, aes(x = hour, y = count, fill = client)) +
  geom_col(position = "stack") +
  facet_wrap(~day_of_week, nrow = 1) +
  labs(x = "Hour of the Day", y = "Count") +
  scale_fill_manual(values = c("blue", "orange"), labels = c("Regular", "Casual")) +
  theme_minimal()


```
```{r}
comment1 <- "Hourly Usage: The graph shows that bike rentals tend to follow a two-peak trend throughout the day, with peaks around 8–9 a.m. and 5–6 p.m., when most people are going to work or school. These peaks show that the bikes are used a lot during times when people usually commute.
Weekdays vs. weekends: On weekdays (Monday through Friday), there are more rentals during the morning and evening rush hours, which is when people usually go to and from work. On the other hand, the rental counts on weekends seem to be spread out more evenly throughout the day, which suggests a more relaxed and not-related-to-commuting usage trend.
Client Type: The color of the client shows if they are a regular or not. It can be seen that both regular and occasional clients use the site in ways that match the general trends listed above. Regular customers who have paid a yearly membership fee tend to use the bike rentals more during rush hour on weekdays. Casual customers who have paid for a five-day membership, on the other hand, use the bike rentals more evenly throughout the week.
These trends show how much people want to rent bikes at different times and for different kinds of customers. They can help the bike-sharing system optimize bike spread, manage station capacity, and figure out what users want.
"
cat(comment1)
```

My comments here.....


## Question 4: How Far? 


Our goal is to find the distance between each pair of stations. You know the position from the lat and long variables in `Stations`. This is enough information to find the distance. The calculation has been implemented in the `haversine()` function. We can import the `haversine()` function using the following command. 

```{r}
source("https://mdbeckman.github.io/dcSupplement/R/haversine.R")
```


`haversine()` is a transformation function. To use it, create a data table where a case is a pair of stations and there are variables for the latitude and longitude of the starting station and the ending station. To do this, join the Station data to itself. The following statements show how to create appropriately named variables for joining.


```{r}
Simple <-
  Stations %>%
  select(name, lat, long) %>%
  rename(sstation = name)

Simple2 <-
  Simple %>%
  rename(estation = sstation, lat2 = lat, long2 = long)

```

Look at the `head()` of `Simple` and `Simple2` and make sure you understand how they are related to `Stations.`


The joining of `Simple` and `Simple2` should match every station to every other station. Since a ride can start and end at the same station, it also makes sense to match each station to itself. This sort of matching does not make use of any matching variables; everything is matched to everything else. This is called a **full outer join**. See Lecture L09 to read more about joins. Note, we can also use the `merge()` function for full outer joins. See the help file for more details.  


We perform the full outer join and then use `haversine()` to compute the distance between each pair of stations.


```{r}
#full outer join
StationPairs <- merge(Simple, Simple2, by = NULL)

#calculating distances
PairDistances <-
  StationPairs %>%
  mutate(distance = haversine(lat, long, lat2, long2)) %>%
  select(sstation, estation, distance)

```


Once you have `PairDistances`, you can join it with `Trips` to calculate the start-to-end distance of each trip. An `inner_join()` is appropriate for finding the distance of each ride.

Display the distribution of the ride distances of the rides. Compare it to the distances between pairs of stations. Are they similar? Why or why not?

Our goal here is to recreate [Figure 18.5](https://dtkaplan.github.io/DataComputingEbook/project-bicycle-sharing.html#fig:bike-hour-count). For this assignment, it is sufficient to make each density on its own plot. There is 1 point of extra credit if you can make the two densities on the same plot. 



```{r}
StationPairs <- merge(Simple, Simple2, by = NULL)


```
```{r}
PairDistances <- StationPairs %>%
  mutate(distance = haversine(lat, long, lat2, long2)) %>%
  select(sstation, estation, distance)

```
```{r}
TripsWithDistances <- inner_join(Trips, PairDistances, by = c("sstation", "estation"))
ggplot(TripsWithDistances, aes(x = distance)) +
  geom_density(fill = "gray", color = "black") +
  labs(x = "Ride Distance (km)", y = "Density")

```




#### Extra Credit: Joining and Variable Names in R

Look at the variables in `Stations` and `Trips` and explain why `Simple` and `Simple2` were given different variable names for the station. 

Optional Extra credit answer here.....





